Composable Generalized Consensus

### 2015-05-15 paper/paper.tex
Work on the description of the Broadcast and Simple-Leader algorithms.

SMR is often implemented using a sequence of instances of Consensus.
There are numerous examples of consensus algorithms in the literature.
However, Abortable Chain Agreement is based on Generalized Consensus. 
In Generalized Consensus \cite{Lamport05GeneralizeConsensus}, the learners of a
SMR algorithm repeatedly learn new values belonging to a \emph{C-struct set}. A
C-struct can be seen as a data structure encoding  sequences of commands modulo
the ordering of commands that commute. The advantage of Generalized
Consensus over Consensus is that
two servers which receive commuting commands in different orders can still reach
agreement on C-structs without further synchronization. 
Our Abortable Chain Agreement can also use C-structs, with the same advantages
as Generalized Consensus. For simplicity, the presentation in this
paper uses the more concrete notion of \emph{command history} (see
\cref{sec:histories}).

Generalized Lattice Agreement \cite{FalerioETAL12GeneralizedLatticeAgreement} allows more efficient implementations of SMR when
commands are partitioned into read-only commands and write-only commands.
Abortable Chain Agreement supports the same optimizations as Generalized Lattice Agreement.
Notably, read-only commands are treated specially so as to optimize their
executions (see \cref{sec:readonly}).

Vertical Paxos algorithms 
\cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication} are SMR algorithms
whose set of replica servers can be reconfigured without stopping to process new
commands. Crashed servers can therefore be replaced by new servers, ensuring the
long-term availability of a replicated service.
Vertical Paxos algorithms use an external reconfiguration master, itself
implemented as a replicated state machine, which determines the servers that should
be removed from or added to the configuration.
ACA shares some aspects of Vertical Paxos algorithms: the task of the
reconfiguration master is similar to the task of the scheduling policy
component of ACA (see \cref{sec:aca}).
However, in ACA, changing round can be used not only to replace crashed servers
but also to change the algorithm that the servers are running. 
Changing algorithm is not possible in Vertical Paxos.
In both Vertical Paxos and ACA, a reliable reconfiguration master, resp. a
reliable scheduling policy, allows tolerating $f$ crashes with $f+1$ servers.

The Abstract Framework \cite{GuerraouiETAL10Next700BftProtocols} proposes to
build Byzantine Fault-Tolerant algorithms as the succession of abortable rounds
called Abstract instances. Abstract instances must satisfy the Abstract
correctness properties. By construction, the composition of any number of
Abstract instances is a correct SMR implementation. However, the Abstract
Framework uses totally ordered sequence of commands, making difficult the
optimization of the execution of commuting and read-only commands that are
possible with ACA.  Moreover, the Abstract Framework does not identify the
crucial role of the scheduling policy for the long-term resilience of a service.
Abortable Chain Agreement combines the ideas of the Abstract Framework,
Generalized Paxos, Generalized Lattice Agreement, and Vertical Paxos in a
unified abstraction.

Finally, the Speculative Linearizability framework
\cite{GuerraouiKuncakLosa12SpeculativeLinearizability} was a first
attempt by the authors of the present paper at generalizing the
Abstract Framework and mechanically proving its properties with the
Isabelle/HOL interactive
proofs assistant.  Like the Abstract Framework, the
Speculative Linearizability framework is based on totally ordered
sequences of commands, with the same drawback for the execution of
commuting and read-only commands. Moreover, Abortable Chain Agreement is both
conceptually simpler and subsumes the Speculative Linearizability
property.  Like Speculative Linearizability, Abortable Chain Agreement
and its main property, the composition theorem (\cref{thm:comp}), are
formalized in Isabelle/HOL (see \cref{sec:isaproofs}).

Abortable Chain Agreement is the most complete framework to date for building
adaptive replication algorithms out of heterogeneous rounds, making state of the
art optimization techniques available to algorithm designers, which was not
possible before.

@@

\section{Abortable Chain Agreement}
\label{sec:aca}

We now consider implementing Chain Agreement as a sequence of \emph{Abortable Chain
Agreement rounds}, where each round makes progress ordering requests or aborts
its execution and passes the baton to the next round in the sequence.
An algorithm composed of a sequence of CGC instances is called an \emph{ACA
algorithm}.

On top of the traditional roles of proposer, acceptor, and learner, ACA
introduces the new role of \emph{switcher} and a new component, the
\emph{scheduling policy}. In ACA, the assignment of the roles of proposer,
acceptor, learner, and switcher can change in each rounds, allowing for
reconfiguration. 
In normal circumstances, the proposers, acceptors,
and learners of a CGC instance work together to implement Chain Agreement.
However, if the environment becomes unfavorable to a running CGC instance, the
scheduling policy can instruct the round to abort its execution and pass the
baton, using the switchers, to a new round.

@@

The scheduling policy determines when to abort a CGC instance and which new ACA
round should take over. The scheduling policy must inform the switchers about the
identity of the next round and must guarantee that all switchers receive the
same information.  In this paper we abstract over the scheduling policy, assuming
that a round that aborts passes the baton to a nondeterministically chosen new
round. A reliable scheduling-policy component can realistically be implemented
as a replicated state machine running on a large number of servers otherwise
dedicated to other tasks
\cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication}. We will see
 in \cref{sec:rounds} that, as in Vertical Paxos, a reliable scheduling policy component allows
tolerating $f$ faults with $f+1$ replicas, while maintaining the long-term
resilience of the service by replacing faulty servers upon round change.

The clients of an ACA algorithm can send their commands to the proposers of any round (in
practice, clients send their commands to proposers of the largest
active round) and servers can use sequences learned in any round
to determine outputs.

ACA specifies the allowed behaviors of the proposers, learners, and switchers
\emph{of a single round}. Similarly to
Chain Agreement, we use the state variables $crashed$, $proposed$,
and, for every learner $l$, $learned\left[ l \right]$.  Additionally,
we use the state variables $inits$, containing the abort sequences
received from the previous round, and $aborts$, containing the abort
sequences transfered to the next round.
Two consecutive CGC instances are composed by letting the $aborts$ variable of the
first round be the $inits$ variable of the second round.

Initially, $crashed$, $proposed$, for every
learner $l$, $learned\left[ l \right]$, $inits$, and $aborts$ are all empty sets.
The next-state relation consists of five types of transitions. The
first three, $Crash\left( x \right)$, $Propose\left( l, c
\right)$, and $Learn\left( l, cs \right)$, for a learner, proposer, or
switcher $x$, for a learner $l$, and for a sequence of commands
$cs$, have the same effect as in Chain Agreement.
ACA introduces two new transitions, $Init\left(
s, cs \right)$ and $Abort\left( s, cs \right)$, for a switcher $s$ and a sequence of
commands $cs$. The transition $Init\left( s, cs \right)$ adds $cs$ to the set $inits$, and models a switcher receiving the
abort sequence $cs$ from the preceding round.
The transition $Abort\left( s, cs \right)$ adds $cs$ to the set $aborts$ and models a switcher transferring the abort sequence $cs$ to a switcher of the next round.
All the transition are guarded by the following conditions:
\begin{compactitem}
    \item[\textbf{Agreement}:] The set $\aset{learned'\left[ l
      \right] : l \in Learners}$ is compatible.
    \item[\textbf{Validity}:] Every abort history or learned history 
        is of the form $h \star cs$, where $cs$ contains only commands
        of the set $proposed$ and $h$ is a history belonging to $inits$.
    \item[\textbf{Initialization}:] If $inits$ is empty, then no
        sequence can be learned and no switcher can abort.
    \item[\textbf{Irrevocability}:] In the case of a $Learn\left( l,h
        \right)$ transition, the new learned history $h$ is an 
        extension of $learned\left[ l \right]$.
    \item[\textbf{Safe Abort}:] Every abort sequence is an extension in the 
        partial order $\leq$ of a learned history.
\end{compactitem}
Note that Agreement and Irrevocability are the same as in Chain
Agreement.

Finally, for every learner $l$, we add a weak fairness condition on the
following state transition: 
\begin{equation}
    \exists  s \in Switchers, cs \in Sequence\left( C \right) : Learn\left( l, cs \right) \vee Abort\left( s,cs \right)
\end{equation}
The weak fairness conditions ensure that when a correct learner has not
learned a sequence containing all commands proposed in the round, then
it eventually learns a new command or, if there is at least one correct switcher, at least one switch value is transfered to the next round. 
Therefore, if a client has submitted a command to a proposer, eventually,
either the client gets a response from the current round, or
the next round is initialized with an abort sequence and the
command can be resubmitted to the next round.

@@

@@

\begin{thm}[Composition Theorem]
    \label{thm:comp}
    In every execution of a succession of CGC instances, the sequence of valuations
    of the variable $proposed$ and, for every learner $l$, of the variables
    $learned\left[ l \right]$ satisfies Chain Agreement.
\end{thm}
\Cref{thm:comp} can be shown by induction on the number of rounds,
using the following lemma.
\begin{lem}
  \label{lem:comp}
    Consider two rounds numbered $i$ and $i+1$. In every execution of a
    succession of CGC instances, the sequence of valuations of the variables $inits
    = inits_i$, $propose = proposed_i \cup propose_{i+1}$, for every learner
    $l$, $learned\left[ l \right] = learned_i\left[ l \right] \cup
    learned_{i+1}\left[ l \right]$and $aborts = aborts_{i+1}$ satisfies ACA.
\end{lem}
\begin{proof}
  A mechanically-checked proof in Isabelle/HOL appears in \cref{sec:isaproofs}.
  The proofs uses a version of the specification of ACA that has a more
  executable form that presented in this section. Moreover,
  instead of sequences of commands, the Isabelle/HOL specification uses the more
  general notion of command history, presented in the next subsection.
  The Isabelle/HOL formalization of ACA uses the theory of I/O automata, which
  is also formalized in the Isabelle/HOL. The proof uses a refinement
  mapping from the composition of two consecutive ACA instances to a single ACA
  instance.
\end{proof}

\Cref{thm:comp} guarantees that the successive composition of any
number of CGC instances implements Chain Agreement.
Therefore, with ACA, one can easily build replicated services that
dynamically adapt to the current operating conditions of the system: a
replicated service has several types of rounds at its disposal, each being
tailored to particular conditions. Each time the conditions change,
the current CGC instance aborts and passes the baton to a more 
appropriate type of round, as determined by the scheduling policy. Moreover, a new type
of CGC instance can always be devised after the system is deployed and added on the fly to
the existing types of rounds.

We now describe two important practical optimizations of ACA.

@@

Chain Agreement can be further optimized by treating 
read-only commands specially, as in Generalized Lattice Agreement
\cite{FalerioETAL12GeneralizedLatticeAgreement}.

@@

The FLP impossibility result \cite{FischerLynchPaterson83ImpossibilityDistributedConsensusOneFaultyProcess}
implies that no algorithm can implement SMR to the letter in practice.
Instead, SMR algorithms guarantee safety (i.e. replica execute the
same sequence of commands) but may not be able to make progress (i.e.
order new commands). To make progress as likely as possible, most SMR
algorithms are structured in consecutive rounds (also called ballots,
views, or epochs). In each round, the algorithm tries to make progress
but cannot be guaranteed to succeed. When the algorithm detects that
progress is no more possible, a new round is started.  The new round
is initialized so as to preserve safety, usually by one of the
acceptors. For example, in Paxos, the leader of a newly started ballot
numbered $i$ determines a value that is \emph{safe at $i$} and
proposes it as first value to the other acceptors of the ballot $i$.
In existing SMR algorithms, new rounds are initialized ad-hoc, which
prevents mixing rounds of two different algorithms.  Instead,
ACA is an abstraction of rounds which establishes a common interface and contract for switching from
one round to the next, enabling switching from any CGC instance to any other.

>>>>

SMR is often implemented using a sequence of instances of Consensus.
There are numerous examples of consensus algorithms in the literature.
However, Composable Generalized Consensus is based on Generalized Consensus. 
In Generalized Consensus \cite{Lamport05GeneralizeConsensus}, the learners of a
SMR algorithm repeatedly learn new values belonging to a \emph{C-struct set}. A
C-struct can be seen as a data structure encoding  sequences of commands modulo
the ordering of commands that commute. The advantage of Generalized
Consensus over Consensus is that
two servers which receive commuting commands in different orders can still reach
agreement on C-structs without further synchronization. 
Our Composable Generalized Consensus can also use C-structs, with the same advantages
as Generalized Consensus. For simplicity, the presentation in this
paper uses the more concrete notion of \emph{command history} (see
\cref{sec:histories}).

Generalized Lattice Agreement \cite{FalerioETAL12GeneralizedLatticeAgreement} allows more efficient implementations of SMR when
commands are partitioned into read-only commands and write-only commands.
Composable Generalized Consensus supports the same optimizations as Generalized Lattice Agreement.
Notably, read-only commands are treated specially so as to optimize their
executions (see \cref{sec:readonly}).

Vertical Paxos algorithms 
\cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication} are SMR algorithms
whose set of replica servers can be reconfigured without stopping to process new
commands. Crashed servers can therefore be replaced by new servers, ensuring the
long-term availability of a replicated service.
Vertical Paxos algorithms use an external reconfiguration master, itself
implemented as a replicated state machine, which determines the servers that should
be removed from or added to the configuration.
CGC shares some aspects of Vertical Paxos algorithms: the task of the
reconfiguration master is similar to the task of the scheduling policy
component of CGC (see \cref{sec:aca}).
However, in CGC, changing round can be used not only to replace crashed servers
but also to change the algorithm that the servers are running. 
Changing algorithm is not possible in Vertical Paxos.
In both Vertical Paxos and CGC, a reliable reconfiguration master, resp. a
reliable scheduling policy, allows tolerating $f$ crashes with $f+1$ servers.

The Abstract Framework \cite{GuerraouiETAL10Next700BftProtocols} proposes to
build Byzantine Fault-Tolerant algorithms as the succession of abortable rounds
called Abstract instances. Abstract instances must satisfy the Abstract
correctness properties. By construction, the composition of any number of
Abstract instances is a correct SMR implementation. However, the Abstract
Framework uses totally ordered sequence of commands, making difficult the
optimization of the execution of commuting and read-only commands that are
possible with CGC.  Moreover, the Abstract Framework does not identify the
crucial role of the scheduling policy for the long-term resilience of a service.
Composable Generalized Consensus combines the ideas of the Abstract Framework,
Generalized Paxos, Generalized Lattice Agreement, and Vertical Paxos in a
unified abstraction.

Finally, the Speculative Linearizability framework
\cite{GuerraouiKuncakLosa12SpeculativeLinearizability} was a first
attempt by the authors of the present paper at generalizing the
Abstract Framework and mechanically proving its properties with the
Isabelle/HOL interactive
proofs assistant.  Like the Abstract Framework, the
Speculative Linearizability framework is based on totally ordered
sequences of commands, with the same drawback for the execution of
commuting and read-only commands. Moreover, Composable Generalized Consensus is both
conceptually simpler and subsumes the Speculative Linearizability
property.  Like Speculative Linearizability, Composable Generalized Consensus
and its main property, the composition theorem (\cref{thm:comp}), are
formalized in Isabelle/HOL (see \cref{sec:isaproofs}).

Composable Generalized Consensus is the most complete framework to date for building
adaptive replication algorithms out of heterogeneous rounds, making state of the
art optimization techniques available to algorithm designers, which was not
possible before.

@@

Before formally introducing our framework, let us see how we can build a modular implementation similar to Fast Paxos by first designing a fast but fragile modular GC algorithm, then combining it unchanged with a more resilient module, obtaining an algorithm similar to Fast Paxos.

\subsection{The Broadcast Algorithm}

In our first algorithm, Broadcast, learners each maintain a variable $hist\left[ l \right]$ which contain the currently accepted history of learner $l$.
For every learner $l$, $accepted\left[ l \right]$ is initialized to the empty history.
A proposer broadcasts its proposal to all the learners which unconditionally accept it by appending the proposal to their $accepted\left[ l \right]$
history and by broadcasting to all the acceptors their new accepted history, $accepted'\left[ l \right]$.

Each acceptor $l$ maintains a variable $learned\left[ l \right]$, initially the empty history, containing its last learned history. A learner $l$
learns a new history $h$ when $h \geq learned\left[ l \right]$ and $h$ is a lower bound of the set consisting of the maximal accepted history received
from each acceptor.

More precisely, the Broadcast algorithm is described by the transition relation obtained as the disjunction of the
following transitions. We do not model the network explicitly using state variables but instead we just say that servers send, broadcast, or receive messages.

\item A $Propose\left( p,c \right)$ transition, for a proposer $p$ and a command $c$, not guarded, broadcasts the message $\aseq{\quo{prop},c}$ to all
the acceptors.

\item An $Accept\left( a \right)$ transition, for an acceptor $a$, enabled when $a$ can receive a $\aseq{\quo{prop},c}$ message from a proposer,
updates $accepted\left[ a \right]$ to $accepted\left[ a \right]\star \aseq{c}$ and broadcasts the message $\aseq{\quo{accepted}, a, accepted'\left[ a \right]}$ to all the learners
($accepted'\left[ a \right]$ is the new value of the variable $accepted\left[ a \right]$).

\item A $Learn\left( l, h \right)$ transition, for a learner $l$ and a history $h$, enabled when $h \geq learned\left[ l \right]$ and $h$ is a lower
bound of the set $\aset{MaxReceivedFrom\left( a \right) : a \in Acceptors}$, where $MaxReceivedFrom\left( a \right)$ is the maximal accepted history received
from the acceptor $a$.

The algorithm Broadcast cannot learn new histories if two acceptors $a_1$ and $a_2$ have two histories which are not compatible because, since an acceptor can only append new proposals to its history $accepted\left[ l \right]$, all the possible future lower bounds of the set $\aset{MaxReceivedFrom\left( a \right) : a \in Acceptors}$ are bounded above by $GLB\left( \aset{accepted\left[ a_1 \right], accepted\left[ a_2 \right]} \right)$.
This situation occurs when non-commuting commands contend: for example, when two different proposers send two commands $c_1$ and $c_2$ that do not commute, the acceptors $a_1$ accepts $c_1$ before $c_2$, and the acceptor $a_2 \neq a_1$ accepts $c_2$ before $c_1$.

To overcome this deadlock, we will make some modifications to the algorithm that will allow us to start a fresh instance $B_2$, called a round, of the Broadcast algorithm.
To allow switching to a fresh new instance of the Broadcast algorithm, we introduce the new role of switcher, that any server can play additionally from being a proposer, acceptor, or learner.
A switcher will extract an accepted history from the first round and try to initialize all the acceptors of the second round with the extracted history.
If we are luckier than in the first round, there may be no contention between switchers and every acceptor $a$ will start accepting new proposals in the second round based on the same initial history.

We write $B_1$ for the first round of the Broadcast algorithm.
The rounds $B_1$ and $B_2$ each have their own copies of the arrays $accepted$ and $learned$, noted $accepted_1$, $accepted_2$, $learned_1$, and $learned_2$ when the round is not clear from the context. 
We assume that network messages are tagged with a round number and that a message tagged with $i\in\aset{1,2}$ can only be received in $B_i$, creating two isolated logical networks.

In our example with the deadlock, switchers have the role of initializing the new round of $B_2$ so that it can resume the execution and overcome the deadlock.
We modify the Broadcast algorithm by introducing new transitions and modifying the $Accept\left( a \right)$ transition as follows.
We add the new variables $status\left[ a \right]$ for every acceptor $a$, which are initialized to $\quo{idle}$. The purpose of $status\left[ a \right]$ is to prevent the acceptor $a$ from accepting commands until its status changes from $\quo{idle}$ to $\quo{ready}$ and to stop it when its status goes to $\quo{stopped}$.
We also add the new arrays of variables $inits\left[ s \right]$ and $aborts\left[ s \right]$, for every switcher $s$, initialized to the empty history in $B_1$ and to $\quo{none}$ in $B_2$. The purpose of the arrays of variables $aborts$ and $inits$ is to transfer state between $B_1$ and $B_2$. The arrays of variables $aborts$ and $inits$ form what we call the \emph{inter-round} interface.
As for the arrays $accepted$ and $learned$, the rounds $B_1$ and $B_2$ each have their own copy of the arrays of variables $status$, $inits$, and $aborts$, and we adopt add subscripts denoting the round when the round is not clear from the context.

\begin{compactitem}
\item We modify the $Accept\left( a \right)$ transition by enabling it only if $status\left[ a \right] = \quo{ready}$ and by having the acceptor $a$ additionally broadcast its new accepted value to the switchers.
Switchers detect that the algorithm is deadlocked when they have received two messages $\aseq{\quo{accepted}, a_1, h_1}$ and $\aseq{\quo{accepted}, a_2, h_2}$ such that $h_1$ and $h_2$ are incompatible.

\item The $Stop\left( s \right)$ transition, for a switcher $s$, is enabled when the switcher has detected a deadlock or has waited for too long without receiving any message from the acceptors. Its effect is to broadcast a $\aseq{\quo{stop}, s}$ message to the acceptors.

\item The $Stop\left( a \right)$ transition, for an acceptor $a$, is enabled when the acceptor $a$ can receive a $\aseq{\quo{stop}, s}$ message from a switcher $s$.
  Its effect is to update $status\left[ a \right]$ to $\quo{stopped}$ and to send the message $\aseq{\quo{stopped},a,accepted\left[ a \right]}$ to the switcher $s$.

\item The $Abort\left( s \right)$ transition, for a switcher $s$, is enabled when $aborts\left[ s \right] = \quo{none}$ and the switcher has received
  at least one $\aseq{\quo{stopped}, a, h}$ message. Its effect is to set the variable $aborts\left[ s \right]$ to the history $h$ received in the $\quo{stopped}$ message. The history $aborts'\left[ s \right]$ is then called an abort history. Because the switcher has received a $\aseq{\quo{stopped},a,h}$ message from the acceptor $a$, the acceptor $a$ no longer accepts new proposals, therefore no history greater than $h$ can ever be learned anymore. Therefore, an abort history can be seen as a safe starting point for a new round to resume execution.

\item The $Init\left( s \right)$ transition, for a switcher $s$, is enabled when $inits\left[ s \right]=\quo{none}$, and its effect is to pick a history $h$, set $inits\left[ s \right]$ to $h$, and to broadcast the message $\aseq{\quo{init}, h}$ to the acceptors.  The history $inits\left[ s \right]$ is then called an init history.
As described below, in $B_2$ composed with $B_1$, this history is constrained to be an abort history of $B_1$.

 \item The $WakeUp\left( a \right)$ transition, for an acceptor $a$, is enabled when $status\left[ a \right]$ is $\quo{idle}$ and there is at least one switcher $s$ such that $inits\left[ s \right] \neq \quo{none}$. Its effect is to set set $status\left[ a \right]$ to $\quo{ready}$ and $accepted\left[ a \right]$ to $inits\left[ s \right]$.

The two rounds $B_1$ and $B_2$ are composed as follows:
\begin{itemize}
  \item We Require that a transition of the composition be a transition of either specification but not both, except for the $Abort\left( s \right)$ and $Init\left( s \right)$ transitions, which are allowed to take place at the same time. 
  \item We require that $aborts_1 = inits_2$ at all times.
\end{itemize}
The composition models two independent rounds being linked only by the switchers passing the abort histories from $B_1$ to $B_2$, where they become init histories: switchers gather abort histories in $B_1$, then they broadcast them as init histories to the acceptors in $B_2$. The acceptors then initialize their state to a received init history in the $WakeUp\left( a \right)$ transition.
This process can be repeated as often as needed, with new rounds $B_3$, $B_4$, etc., to try to make progress learning new histories.

The Broadcast algorithm can no more learn new histories whenever one acceptor crashes because the learners wait for the lower bound of the accepted histories of all acceptors to increase. However, the round-change mechanism is very resilient to faults and can abort anyway thanks to the timeout triggering the $Stop\left( s \right)$ transition: a single live acceptor with a single live switcher can abort the execution and change round.

To allow a new round to make progress after some acceptors crashed in a previous round, we introduce a component called the switching policy. With the Broadcast rounds, the task of the switching policy is to update the set of acceptors from one round to the next, ensuring that failed acceptors are removed or replaced by live ones. The switching policy must be queried by the switchers upon changing round to determine the new set of acceptors. This information can then be piggybacked to the
acceptors and learners, or the acceptors and learners can directly query the switching policy. 
The set of switchers can be changed from round to round in the same fashion. Moreover, in an algorithm composed of several different types of rounds, the switching policy will also be responsible for choosing the type of round to execute next.

\section{The Simple-Leader Algorithm}

The Broadcast algorithm can process a proposal in two message delays when there is no contention on non-commuting commands. However, if there is
repeated contention on non-commuting commands, the algorithm may not make progress despite changing round, as the acceptors of the each new round
may be initialized with incompatible histories. The Fast Paxos algorithm \cite{Lamport06FastPaxos} is able to learn new histories under high contention by
designating a leader server that will totally order the proposals. However, to make our Broadcast algorithm work like Fast Paxos, we would have to modify it substantially. If we had spent effort testing Broadcast implementations and proving it correct, this effort would be wasted and we would have to redo all the tests and proofs from scratch.

Instead, we will now show how to obtain an algorithm with performance characteristics close to those of Fast Paxos, employing a leader to deal with contention, without modifying at all our existing Broadcast algorithm.
To do so, we define a new kind of rounds, the Simple-Leader rounds, which uses the same inter-round interface as the Broadcast algorithm. 
The inter-round interface will allow use to combine Simple-Leader rounds and Broadcast rounds in any order, obtaining an algorithm which can adapt its strategy to its environment, selecting the Broadcast algorithm under low contention and switching to the more expensive Simple-Leader algorithm when there is contention.

In the Simple-Leader algorithm, the proposers send their proposals to a distinguished acceptor called the leader. 
The leader totally orders the proposals it receives and forwards them to all the acceptors which accept them in the order defined by the leader.
As in the Broadcast algorithm, acceptors broadcast their accepted histories to the learners, which learn a new history when it is a lower bound of 
the set of $\aset{MaxReceivedFrom\left( a \right) : a \in Acceptors}$.
Thanks to its leader, a Simple-Leader round continues to learn new histories even when proposals conflict because the leader ensures that the accepted histories of the acceptors remain consistent.

The Simple-Leader algorithm treats all proposals equally and does not take advantage of commutativity. When the leader does not crash, each proposal is processed in three message delays. In comparison, in the absence of failures, the Broadcast algorithm processes non-conflicting proposals in two message delays but can take arbitrarily long to process conflicting proposals.

To initialize a Simple-Leader round, the switchers send their init histories to the leader, which choose a unique one that it forwards to the acceptors, guaranteeing that all the acceptors are initialized with the same history even when conflicting init histories are sent concurrently.

The Simple-Leader algorithm can not learn new histories if its leader, or any other acceptor, crashes.
To recover from this deadlock we pass the baton to a new round, as in the Broadcast algorithm, with a similar same mechanism: a switcher that times-out waiting for new messages from the acceptors broadcast a stop message, gathers the accepted history from at least one of the acceptors, and aborts.

Let us specify more precisely the Simple-Leader algorithm.

The specification of Simple-Leader algorithm uses the sets   
 $inits$, $aborts$, the arrays of variables $learned\left[ l \right]$,
$status\left[ a \right]$, and $accepted\left[ a \right]$, and two
additional arrays $accepted\left[ l \right]\left[ a \right]$ and
$stop\left[ s \right]\left[ a \right]$, where $l$ is a learner, $a$ is
an acceptor, and $s$ is a switcher.

Initially, the variables $inits$, $aborts$, and, for every
learner $l$, $learned\left[ l \right]$ are empty sets. For every
acceptor $a$, $status\left[ a \right]$ is ``idle'' and $hist\left[ a
\right]$ is the special value $None$. For every learner $l$ and acceptor $a$,
$accepted\left[ l \right]\left[ a \right]$ is the special value
$None$. For every switcher $s$ and acceptor $a$, $stopped\left[ s
\right]\left[ a \right]$ is set to $false$.  There are initially no
messages in the network.

Thanks to the inter-round interface, Simple-Leader rounds and Broadcast rounds can be composed in any order.
To compose a Simple-Leader round $SL_1$ with an Broadcast round $ABA_2$, we compose their specifications as before, adding the constraint that the set $aborts_1$ be equal to the set $inits_2$ at all times.

Informally, the composition is correct because abort histories are always safe in the sense that any of the abort histories could be the next learned history.
However not all can be learned and the initialization mechanism of the rounds ensures that at most one is chosen.
In the next section we present Composable Generalized Consensus, a specification of the behavior of rounds at the level of proposals, learned histories, init histories, and abort histories. Our composition theorem ensures that any set of rounds individually satisfying Composable Generalized Consensus can be composed in any order to yield an algorithm satisfying Composable Generalized Consensus.

\section{Composable Generalized Consensus}
\label{sec:aca}

We now consider implementing Generalized Consensus as a sequence of \emph{Composable Generalized Consensus rounds}, where each round makes progress
ordering requests or aborts its execution and passes the baton to the next round in the sequence. An algorithm composed of a sequence of CGC instances
is called a \emph{CGC algorithm}.

On top of the traditional roles of proposer, acceptor, and learner, CGC
introduces the new role of \emph{switcher} and a new component, the
\emph{scheduling policy}. In CGC, the assignment of the roles of proposer,
acceptor, learner, and switcher can change in each rounds, allowing for
reconfiguration. 
In normal circumstances, the proposers, acceptors,
and learners of a CGC instance work together to implement Generalized Consensus.
However, if the environment becomes unfavorable to a running CGC instance, the
scheduling policy can instruct the round to abort its execution and pass the
baton, using the switchers, to a new round.

@@

The scheduling policy determines when to abort a CGC instance and which new CGC
round should take over. The scheduling policy must inform the switchers about the
identity of the next round and must guarantee that all switchers receive the
same information.  In this paper we abstract over the scheduling policy, assuming
that a round that aborts passes the baton to a nondeterministically chosen new
round. A reliable scheduling-policy component can realistically be implemented
as a replicated state machine running on a large number of servers otherwise
dedicated to other tasks
\cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication}. We will see
 in \cref{sec:rounds} that, as in Vertical Paxos, a reliable scheduling policy component allows
tolerating $f$ faults with $f+1$ replicas, while maintaining the long-term
resilience of the service by replacing faulty servers upon round change.

The clients of a CGC algorithm can send their commands to the proposers of any round (in
practice, clients send their commands to proposers of the largest
active round) and servers can use sequences learned in any round
to determine outputs.

CGC specifies the allowed behaviors of the proposers, learners, and switchers
\emph{of a single round}. Similarly to
Generalized Consensus, we use the state variables $crashed$, $proposed$,
and, for every learner $l$, $learned\left[ l \right]$.  Additionally,
we use the state variables $inits$, containing the abort sequences
received from the previous round, and $aborts$, containing the abort
sequences transfered to the next round.
Two consecutive CGC instances are composed by letting the $aborts$ variable of the
first round be the $inits$ variable of the second round.

Initially, $crashed$, $proposed$, for every
learner $l$, $learned\left[ l \right]$, $inits$, and $aborts$ are all empty sets.
The next-state relation consists of five types of transitions. The
first three, $Crash\left( x \right)$, $Propose\left( l, c
\right)$, and $Learn\left( l, cs \right)$, for a learner, proposer, or
switcher $x$, for a learner $l$, and for a sequence of commands
$cs$, have the same effect as in Generalized Consensus.
CGC introduces two new transitions, $Init\left(
s, cs \right)$ and $Abort\left( s, cs \right)$, for a switcher $s$ and a sequence of
commands $cs$. The transition $Init\left( s, cs \right)$ adds $cs$ to the set $inits$, and models a switcher receiving the
abort sequence $cs$ from the preceding round.
The transition $Abort\left( s, cs \right)$ adds $cs$ to the set $aborts$ and models a switcher transferring the abort sequence $cs$ to a switcher of the next round.
All the transition are guarded by the following conditions:
\begin{compactitem}
    \item[\textbf{Agreement}:] The set $\aset{learned'\left[ l
      \right] : l \in Learners}$ is compatible.
    \item[\textbf{Validity}:] There is a history $h$ belonging to $inits$ such that every abort history or learned history 
        is of the form $h \star cs$, where $cs$ contains only commands
        of the set $proposed$.
    \item[\textbf{Initialization}:] If $inits$ is empty, then no
        sequence can be learned and no switcher can abort.
    \item[\textbf{Irrevocability}:] In the case of a $Learn\left( l,h
        \right)$ transition, the new learned history $h$ is an 
        extension of $learned\left[ l \right]$.
    \item[\textbf{Safe Abort}:] Every abort sequence is an extension in the 
        partial order $\leq$ of a learned history.
\end{compactitem}
Note that Agreement and Irrevocability are the same as in Generalized Consensus.

If a client has submitted a command to a proposer, eventually,
either the client gets a response from the current round, or
the next round is initialized with an abort sequence and the
command can be resubmitted to the next round.

@@

TODO: frame in terms of composability.

@@

\begin{thm}[Composition Theorem]
    \label{thm:comp}
    In every execution of a succession of CGC instances, the sequence of valuations
    of the variable $proposed$ and, for every learner $l$, of the variables
    $learned\left[ l \right]$ satisfies Generalized Consensus.
\end{thm}
\Cref{thm:comp} can be shown by induction on the number of rounds,
using the following lemma.
\begin{lem}
  \label{lem:comp}
    Consider two rounds numbered $i$ and $i+1$. In every execution of a
    succession of CGC instances, the sequence of valuations of the variables $inits
    = inits_i$, $propose = proposed_i \cup propose_{i+1}$, for every learner
    $l$, $learned\left[ l \right] = learned_i\left[ l \right] \cup
    learned_{i+1}\left[ l \right]$and $aborts = aborts_{i+1}$ satisfies CGC.
\end{lem}
\begin{proof}
  A mechanically-checked proof in Isabelle/HOL appears in \cref{sec:isaproofs}.
  The proofs uses a version of the specification of CGC that has a more
  executable form that presented in this section. Moreover,
  instead of sequences of commands, the Isabelle/HOL specification uses the more
  general notion of command history, presented in the next subsection.
  The Isabelle/HOL formalization of CGC uses the theory of I/O automata, which
  is also formalized in the Isabelle/HOL. The proof uses a refinement
  mapping from the composition of two consecutive CGC instances to a single CGC
  instance.
\end{proof}

\Cref{thm:comp} guarantees that the successive composition of any
number of CGC instances implements Generalized Consensus.
Therefore, with CGC, one can easily build replicated services that
dynamically adapt to the current operating conditions of the system: a
replicated service has several types of rounds at its disposal, each being
tailored to particular conditions. Each time the conditions change,
the current CGC instance aborts and passes the baton to a more 
appropriate type of round, as determined by the scheduling policy. Moreover, a new type
of CGC instance can always be devised after the system is deployed and added on the fly to
the existing types of rounds.

We now describe two important practical optimizations of CGC.

@@

Generalized Consensus can be further optimized by treating 
read-only commands specially, as in Generalized Lattice Agreement
\cite{FalerioETAL12GeneralizedLatticeAgreement}.

@@

The FLP impossibility result \cite{FischerLynchPaterson83ImpossibilityDistributedConsensusOneFaultyProcess}
implies that no algorithm can implement SMR to the letter in practice.
Instead, SMR algorithms guarantee safety (i.e. replica execute the
same sequence of commands) but may not be able to make progress (i.e.
order new commands). To make progress as likely as possible, most SMR
algorithms are structured in consecutive rounds (also called ballots,
views, or epochs). In each round, the algorithm tries to make progress
but cannot be guaranteed to succeed. When the algorithm detects that
progress is no more possible, a new round is started.  The new round
is initialized so as to preserve safety, usually by one of the
acceptors. For example, in Paxos, the leader of a newly started ballot
numbered $i$ determines a value that is \emph{safe at $i$} and
proposes it as first value to the other acceptors of the ballot $i$.
In existing SMR algorithms, new rounds are initialized ad-hoc, which
prevents mixing rounds of two different algorithms.  Instead,
CGC is an abstraction of rounds which establishes a common interface and contract for switching from
one round to the next, enabling switching from any CGC instance to any other.
